Improving Item-Item Similarity Estimation during Collaborative Filtering

### 2014-03-25 paper/paper.tex
Tweak Score-based section.

In the following sections, we present two approaches to better approximate the
true similarity given a limited number of common reviewers. Both approaches are
motivated by probabilistic models with distributions grounded in the data. The
first approach leverages the observed similarity score and the number of common
users to calculate the true similarity. The second approach leverages the
rating-vectors themselves.

@@

We model $Y$ as a Normal distribution 
\begin{align}
Y \sim N(\mu, \sigma_{s}^2)
\end{align}
where $\mu$ and $\sigma_{s}^2$ are the average similarity score and the variance 
of similarity scores of all pairs of items, respectively. Since $X_n$ represents
a noisy reading of the true similarity $Y$, we model $X_n | Y=y$ as a Gaussian
error around $y$:
\begin{align}
(X_n | Y=y) \sim N(y, \sigma_{err, n}^2)
\end{align}

Note that $\sigma_{err, n}^2$ represents how noisy $Sim_n$ is as an estimation of
$TrueSim$ and therefore should decrease as $n$ increases. The probability
distribution of $X_n$ can be visualized in Figure 1.

@@

The problem of approximating the $TrueSim$ from $Sim_n$ can now be seen as
finding the value of $Y$ that most likely produced $X_n$. In other words we 
desire the Maximum Likelihood Estimate of $Y | X_n$:
\begin{align}
\hat{y} &= \argmax_yP(Y=y|X_n=x) 
\\&= \argmax_y\left[\frac{P(X_n=x | Y=y)P(Y=y)}{P(X_n=x)}\right]
\\&= \argmax_y\left[P(X_n=x|Y=y)P(Y=y)\right]
\\&= 
\argmax_y\left[\frac{1}{\sigma_{err,n}\sqrt{2\pi}}\exp{\left(\frac{-(x-y)^2}
{2\sigma_{err,n}^2}\right)}
\frac{1}{\sigma_{s}\sqrt{2\pi}}\exp{\left(\frac{-(y-\mu)^2}
{2\sigma_{s}^2}\right)}\right]
\\&= \argmin_y\left[\frac{(x-y)^2}{2\sigma_{err,n}^2} +
\frac{(y-\mu)^2}{2\sigma_{s}^2}\right]
\\&= \argmin_y\left[\left(\sigma_{s}^2+\sigma_{err,n}^2\right)y^2 - 
2\left(\sigma_{s}^2x+\sigma_{err,n}^2\mu\right)y\right]
\end{align}
We find the exact minimum by taking the derivative with respect to $y$ and
setting it to zero:
\begin{align}
\frac{d}{dy}\left[\left(\sigma_{s}^2+\sigma_{err,n}^2\right)y^2 - 
2\left(\sigma_{s}^2x+\sigma_{err,n}^2\mu\right)y\right]
&= 2\left(\sigma_{s}^2+\sigma_{err,n}^2\right)y - 
2\left(\sigma_{s}^2x+\sigma_{err,n}^2\mu\right) 
\\&= 0
\end{align}
This produces the following linear equation for $\hat{y}$ in terms of $x$:
\begin{align}
\hat{y} &= \frac{\sigma_{s}^2x+\sigma_{err,n}^2\mu}{\sigma_{s}^2+\sigma_{err,n}^2}
\end{align}
which can be rewritten as:
\begin{align}
\left(\hat{y} - \mu\right) &= \frac{\sigma_{s}^2}{\sigma_{s}^2+\sigma_{err,n}^2}
\left(x-\mu\right)
\end{align}

The model suggests that there is a linear correlation between $X_n$ ($Sim_n$)
and $Y$ ($TrueSim$).\footnote{It is worth noting that despite
the resemblance to Slope One Predictors \cite{Lemire1998}, this is a fundamentally 
different technique. Here, we have a linear relationship between observed 
scores and true scores, whereas the Slope One model seeks a linear relationship 
between the ratings a user gives to one items versus another.} 
Moreover, the best predicted $TrueSim$ is a linear 
combination of the average true similarity and the observed similiarity. Since $\sigma_{err,n}^2$ is 
nonnegative, the slope is always less than or equal to one, meaning that the 
observed distance from $\mu$ (the average true similarity) overestimates the 
true distance. This is reasonable, since the prior suggests that the true 
similarity is likely to be near $\mu$.

This model requires a different parameter $\sigma_{err,n}^2$ for each $n$.
To gain more generality, we model $\sigma_{err,n}^2$ as a function of $n$
instead of considering them independent of each other. Recall that 
$\sigma_{err,n}^2$ is the noisiness of the observed $Sim_n$ about $TrueSim$. 
Since $TrueSim$ is $Sim$ when all users that will rate the items 
have rated them, $Sim_n$ can be thought of as sampling $n$ users from the 
set used to calculate $TrueSim$. Although $Sim$ is usually more complicated 
than an average, the Central Limit Theorem motivates the intuition that the 
variance of $Sim$ would decrease as the sqaure root of $n$. Thus we model 
$\sigma_{err,n}^2$ as:
\begin{align}
\sigma_{err,n}^2 = \frac{\alpha}{\sqrt{n}},
\end{align}
yielding the single parameter model:
\begin{align}
\left(Y - \mu\right) = \frac{\sigma_{s}^2}{\sigma_{s}^2+\frac{\alpha}{\sqrt{n}}}
\left(X_n-\mu\right)
\end{align}

\subsection*{Technique}
While it may be impossible to get $TrueSim$ for any pair, we
can get close by calculating $Sim_n$ when a large number of users have rated
both items. Formally, we define $GoldSim$ = $Sim_N$ where $N$ is a number 
large enough that $GoldSim \approx TrueSim$. Using this approach to approximate
$TrueSim$, we can look at the distribution of $GoldSim$ scores (Fig. 2) and see 
that the Normal distribution of $TrueSim$ is a reasonable model.

>>>>

In the following sections, we present two approaches to better approximate the
true similarity given a limited number of common reviewers. Both approaches are
motivated by probabilistic models with distributions grounded in the data. The
first method leverages the observed naive similarity estimate and the number of
common users to better estimate the true similarity. The second method leverages
the observed ratings themselves for the same ends.

@@

We model $Y$ as a Normal distribution 
\begin{align}
Y \sim N(\mu, \sigma_{s}^2),
\end{align}
where $\mu$ and $\sigma_{s}^2$ are the average similarity score and the variance
of similarity scores of all pairs of items, respectively. Since $X_n$ represents
a noisy reading of the true similarity $Y$, we model $X_n | Y=y$ as a Gaussian
error around $y$:
\begin{align}
(X_n | Y=y) \sim N(y, \sigma_{err, n}^2)
\end{align}
Note that $\sigma_{err, n}^2$ represents how noisy $Sim_n$ is as an estimate of
$TrueSim$ and therefore should decrease as $n$ increases. The probability
distribution of $X_n$ is illustrated in Fig. 1.

@@

The problem of approximating $TrueSim$ given $Sim_n$ can now be seen as finding
the value of $Y$ that most likely produced $X_n$. In other words we desire the
Maximum Likelihood Estimate of $Y | X_n$:
\begin{align}
\hat{y} &= \argmax_yP(Y=y|X_n=x) 
\\&= \argmax_y\left[\frac{P(X_n=x | Y=y)P(Y=y)}{P(X_n=x)}\right]
\\&= \argmax_y\left[P(X_n=x|Y=y)P(Y=y)\right]
\\&= 
\argmax_y\left[\frac{1}{\sigma_{err,n}\sqrt{2\pi}}\exp{\left(\frac{-(x-y)^2}
{2\sigma_{err,n}^2}\right)}
\frac{1}{\sigma_{s}\sqrt{2\pi}}\exp{\left(\frac{-(y-\mu)^2}
{2\sigma_{s}^2}\right)}\right]
\\&= \argmin_y\left[\frac{(x-y)^2}{2\sigma_{err,n}^2} +
\frac{(y-\mu)^2}{2\sigma_{s}^2}\right]
\\&= \argmin_y\left[\left(\sigma_{s}^2+\sigma_{err,n}^2\right)y^2 - 
2\left(\sigma_{s}^2x+\sigma_{err,n}^2\mu\right)y\right]
\end{align}
We find the exact minimum by taking the derivative with respect to $y$ and
setting it to zero:
\begin{align}
\frac{d}{dy}\left[\left(\sigma_{s}^2+\sigma_{err,n}^2\right)y^2 - 
2\left(\sigma_{s}^2x+\sigma_{err,n}^2\mu\right)y\right]
&= 2\left(\sigma_{s}^2+\sigma_{err,n}^2\right)y - 
2\left(\sigma_{s}^2x+\sigma_{err,n}^2\mu\right) 
\\&= 0
\end{align}
This produces the following linear equation for $\hat{y}$ in terms of $x$
\begin{align}
\hat{y} &=
\frac{\sigma_{s}^2x+\sigma_{err,n}^2\mu}{\sigma_{s}^2+\sigma_{err,n}^2},
\end{align}
which can be rewritten as
\begin{align}
\left(\hat{y} - \mu\right) &= \frac{\sigma_{s}^2}{\sigma_{s}^2+\sigma_{err,n}^2}
\left(x-\mu\right).
\end{align}

The model suggests that there is a linear correlation between $X_n$ ($Sim_n$)
and $Y$ ($TrueSim$).\footnote{It is worth noting that despite the resemblance to
Slope One Predictors \cite{Lemire1998}, this is a fundamentally different
technique. Here, we have a linear relationship between observed scores and true
scores, whereas the Slope One model seeks a linear relationship between the
ratings a user gives to different items.} Moreover, the best predicted $TrueSim$
is a linear combination of the average true similarity and the observed
similiarity. Since $\sigma_{err,n}^2$ is nonnegative, the slope is always less
than or equal to one, meaning that, on average, the observed distance from $\mu$
overestimates the true distance from $\mu$. This is reasonable, since the prior
distribution suggests that the true similarity is likely to be near $\mu$.

This model requires a different parameter $\sigma_{err,n}^2$ for each $n$.  To
gain more generality, we model $\sigma_{err,n}^2$ as a function of $n$ instead
of considering them independent of each other. Recall that $\sigma_{err,n}^2$ is
the noisiness of the observed $Sim_n$ about $TrueSim$.  $Sim$ converges to
$TrueSim$ as the number of users $n$ increases.  The Central Limit Theorem
therefore motivates the intuition that the variance of $Sim$ would decrease as
the square root of $n$. Thus we model $\sigma_{err,n}^2$ as
\begin{align}
\sigma_{err,n}^2 = \frac{\alpha}{\sqrt{n}},
\end{align}
yielding the single parameter model
\begin{align}
\left(Y - \mu\right) = \frac{\sigma_{s}^2}{\sigma_{s}^2+\frac{\alpha}{\sqrt{n}}}
\left(X_n-\mu\right).
\end{align}

\subsection*{Technique}
To evaluate this method of improving similarity estimates, we would like to know
{\em a priori} the true similarity for a set of item pairs. This way, we could
compute the error in estimates made based on only partial rating data. Of
course, the metric $TrueSim$ is a theoretical construct and not something we
have access to in the real world. Fortunately, we can achieve good
approximations of $TrueSim$ given sufficient data. We therefore define $GoldSim$
= $Sim_N$ where $N$ is a number large enough that $GoldSim \approx TrueSim$.
Using this approach to approximate $TrueSim$, we can look at the distribution of
$GoldSim$ scores (Fig. 2) and see that the Normal distribution of $TrueSim$ is a
reasonable model.
